Human Factors Influencing Visual Statistical Inference

### 2015-08-28 sociology_paper.Rnw
intro

The lineup protocol introduced in \citet{buja:2009} can be used to quantify the significance of graphical findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and provide methods to obtain the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called the actual data plot, is placed in a layout of null plots. The null plots  are generated from the model specified by a null hypothesis. Commonly, this null hypothesis assumes independence/no structure in the data. An observer is then asked to evaluate the lineup. If the  actual data plot is detected by the observer, this counts as evidence against the null hypothesis. Based on the number of evaluations by independent observers and the number of times that the data plot is identified, we are then able to evaluate the significance of the graphical relationship in the same way and the same rigor as a conventional hypothesis test. % is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.

>>>>

The lineup protocol introduced in \citet{buja:2009} can be used to quantify the significance of graphical findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of their major contributions   is to define the power of  visual tests and to provide a method for obtaining the power for a particular lineup. In some scenarios, the power of a visual test  was seen to outperform that of a conventional test.

In visual inference, the test statistic is a plot of the observed data, called the {\it (actual) data plot}. To create a lineup the data plot is placed randomly in a grid of null plots. Null plots  are rendered from data generated as  specified by a null hypothesis. Often, this null hypothesis assumes independence or no structure in the data. An observer is then asked to evaluate the lineup. If the  actual data plot is detected by the observer, this counts as evidence against the null hypothesis. Based on the number of evaluations by independent observers and the number of times that the data plot is identified, we are  able to evaluate the significance of the graphical relationship in the same way and the same rigor as a conventional hypothesis test. % is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.
